An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2015-01-13 paper.tex
Finished edits.

In the model selection example from Section \ref{sec:efficiency}, the
models have two-dimensional parameter spaces; in \cite{Farr2010} the
highest-dimensional model had five dimensions.  As the number of
dimensions increases the interpolation becomes more difficult for two
reasons.  First, the number of samples required for a given
number of subdivisions in each dimension grows exponentially with
the number of dimensions; hence, for a reasonable number of samples, a high-dimensional kD tree will have few subdivisions along each dimension. \dan{cleaned the syntax up a little in the preceding.} Second, as
the dimensionality increases the volume ``at the \ilya{edges /} corners'' of each kD
tree box becomes more significant; this problem is even more
pronounced when the distribution of samples does not align with the
coordinate axes along which the tree subdivides.  In this section, we
discuss practical solutions to this issue, \dan{illustrating our solutions with a single-model MCMC. We conclude with an example of a successful
application to gravitational-wave parameter estimation.}

\ilya{[There is at least one paragraph missing here.  What's this "single-model MCMC" and why does it require PDF interpolation?  Are you running several MCMCs in succession as increasingly good approximations to the desired posterior?  Are you using higher-temperature chains to provide a jump proposal guess for lower temperatures?   More details needed.]} \dan{[I added a brief explanation of a single-model MCMC at its earliest mention in the introduction. Discussion of the single-model MCMC and the application of our kD-interpolation proposal seems like it would be better discussed after we introduce our proposal, no? Will, do you remember how we obtained the original samples that we fed into the tree?]}

\dan{A naive jump proposal based on the kD tree structure would select one sample stored in the tree at random and perform the binary tree search given in Section \ref{sec:kDTree} to find that sample's neighborhood. A new point is then proposed uniformly from within this box. A better proposal however, would account for covariances between parameters in the parameter space; the naive proposal mentioned above, drawing uniformly from within a box, would yield a low acceptance ratio if parameters are heavily correlated. To address significant covariance between samples, the modified jump proposal does not descend to the leaf-nodes of the tree, but instead stops descending whenever the number of samples in the current box, $N_{box}$, falls below some threshold, $N_{crit}$. We then find the principal axes of the samples
contained in this box by calculating the eigenvectors of the covariance matrix of these samples. We use these eignevectors to draw a new box. This box -- henceforth called a covariance cell -- contains all $N_{box}$ samples from the kD tree box, is centered on the mean of these samples, and has edges that are aligned with the principal axes. Furthermore, this box is drawn around the samples as tightly as possible; in other words, for each edge, there exists a point in the covariance cell that lies along that edge. Figure
\ref{fig:PCC} illustrates these two boxes.}

\begin{figure}
  \includegraphics[width=0.8\columnwidth]{PCC}
  \caption{\label{fig:PCC} A two-dimensional illustration of the two
    boxes involved in the modified kD proposal.  The larger box
    aligned with the axes is the normal kD-tree box containing the
    given samples, while the tighter box is the box from which the
    modified kd-interpolated jump proposal draws a point. }
\end{figure}

\dan{The jump proposal chooses a stored sample at random and finds the largest kD tree box containing it and fewer than $N_{crit}$ total samples. It the draws the covariance cell and proposes a new set of parameters uniformly from the intersection of the covariance cell with the original kD tree box.}
The jump probability for this jump proposal is given by
%
\bel{modforward} P(\vtheta_i \rightarrow \vtheta_{i+1}) =
\frac{\Nbox}{N V} , \ee
%
\dan{where N is the total number of samples in the kD-tree, $V$ is the
volume of the intersection between the kD tree box and the correlation cell containing $\vtheta_{i+1}$. When used as the only jump proposal for an
RJMCMC, it is important that this proposal be capable of
proposing points in all allowed regions of parameter space; on the
other hand, when the kD proposal is used in a suite of other jump proposals, this is no longer a requirement.}

@@

\dan{This technique trivially violates detailed balance since the jump proposal distribution depends on the past history (the number of samples in the kD tree box and in the whole tree). One way to address this problem requires continuously accumulating samples into the kD tree as the MCMC explores more
of the parameter space. As the number of samples drawn from the posterior PDF increases, the neighborhoods around each sample decrease with increasing sample density and the calculated covariances between samples in the kD tree better approximate the true covariances between parameters in the equilibrium posterior PDF. ence, for large $N$, the change in $P(\vtheta^{(j)} \rightarrow \vtheta^{(j+1)})$ approaches zero and the jump proposal does a progressively better job of sampling the equilibrium posterior [someone confirm or deny!]}  \ilya{[See above -- it's not clear what exactly we are doing here in general.]}
%Because this introduces a dependence on the
%previous states of the MCMC, the jump proposal is no longer
%Markovian.  However, it remains \emph{asymptotically} Markovian \ilya{[And being asymptotically Markovian (whatever that actually means -- at least a reference?) -- is provably sufficient to guarantee that we sample from the equilibrium distribution even though we nominally violate detailed balance? (again, a reference?)]} , which
%ensures that the equilibrium distribution of the MCMC is correct.
To efficiently insert samples into the tree as the chain accumulates them,
the algorithm from Section \ref{sec:kDTree} must be modified:
\begin{enumerate}
\item Instead of partitioning the sample set at its median, we now
  partition the bounding box at its geometric center along a
  particular dimension, which cycles as we descend the tree levels.
  Note that this allows for empty boxes if all the samples cluster to
  one side of the domain.
\item When inserting a sample, we descend the tree to a leaf node,
  which now can be either empty or contain one sample.  If it is empty,
  we insert the sample at this node.  If it contains one sample, we
  subdivide the appropriate dimension, and place both samples into the
  two sub-boxes.  If both land in the same sub-box, we continue
  subdividing, until each box in the sub-sub-sub... tree contains one
  or zero samples.
\end{enumerate}
Insertion of a sample is $\order{\log N}$, since it involves a constant
amount of work per tree level, and there are $\log N$ tree levels.  \ilya{[No -- the number of tree levels can be arbitrarily large if the original samples are clustered into a very small region of the prior volume...]} \dan{[Is this true? Couldn't it be arbitrarily small, depending on how far into the tree you want to burrow (i.e. depending on $N_{crit}$)?]}
Neighborhood lookup proceeds as before, and is therefore also
$\order{\log N}$.  \ilya{[Perhaps $\order{\log N + \Nbox}$, since must recompile the inner bounding box every time -- and, depending on problem, $\Nbox$ could be $\gg \log N$?]} \dan{[Need only recompute the covariance cell if a new sample was accepted into the corresponding kD tree cell between calls to that kD tree cell. I'm not sure how that would factor in here, but I feel like it should.] }

We have implemented this modified kD tree proposal as one of many jump
proposals in the \texttt{LALInferenceMCMC} sampler
\cite{vanderSluys:2008a,Raymond:2010,Raymond2012}.
\texttt{LALInferenceMCMC} is a MCMC code \cite{lalinf_method}, based on the LIGO algorithms
library, designed to sample the posterior on parameters of \ilya{merging}
compact-object binaries (masses, orbital orientation, distance, sky
position, etc) \ilya{encoded in} their gravitational wave \ilya{signatures} in
Earth-based gravitational wave detectors like LIGO.  The simplest such
signal has a nine-dimensional parameter space, and the posterior often
includes near-degeneracies and multiple modes that make convergence of
the MCMC chain very slow with traditional proposals.  Figure
\ref{fig:accratio} shows the acceptance ratio for the kD-tree jump
proposal in such a simulation.  \dan{For this case, the kD-tree jump proposal is one of several jump proposals used during the MCMC, and the kD tree itself is updated with accepted samples whenever its jump proposal is called.}
In spite of the very low acceptance
rate of the proposal, applying the kD-tree proposal to one out of
\dan{20} proposed jumps improved the convergence time of the simulation
by a factor of two compared to the standard suite of proposals because
it is particularly efficient at producing ``mode-hopping'' jumps,
which are difficult to produce with the other proposals in
\texttt{LALInferenceMCMC}.  \ilya{[Again, what exactly is reported in this figure?  Results with a jump proposal based on an interpolated PDF (of what)?  A proposal that's being updated on the fly as more samples come in?]}

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{linearlog3}
  \end{center}
  \caption{\label{fig:accratio} The cumulative parameter acceptance
    ratio for the modified kD-\dan{tree}jump proposal used in the
    \texttt{LALInferenceMCMC} code as a function of the number of
    steps (in hundreds). In spite of the small acceptance ratio, the
    kD tree proposal, applied to one in \dan{20} proposed jumps,
    improved the convergence time of the sampler by a factor of two
    compared to using only the standard suite of proposals. \dan{[Taking a leap here, someone please confirm/deny/add qualifiers] The initial decrease in acceptances arises from insufficient population of the kD tree with too few samples, while the leveling-off later in the MCMC indicates convergence.} \ilya{[I see structure in this figure (a high acceptance probability, followed by a dip, followed by an asymptote).  Is this a fluke, or do you always see this -- and, if so, why?  BTW, what's the autocorrelation length (just to give me a sense of scale, so I am not reading too much into meaningless fluctuations within an AC length)?]} \dan{[Don't recall the autocorrelation length. Will?]}}
\end{figure}

@@

The need to compare evidences for multiple models arises in a large
variety of physical and astronomical contexts.  In this paper, we
described a technique that allows for efficient evidence
computations via a Reversible-Jump Markov chain Monte Carlo.  This
technique solves the usual problem of finding good inter-model jump
proposals in an RJMCMC by using a kD tree to quickly and accurately
interpolate an approximate posterior PDF from a single-model MCMC run,
and then proposing efficient inter-model
jumps \ilya{from} this interpolated PDF.

We demonstrated the efficiency of this technique on a toy
model-comparison problem described in Section \ref{sec:efficiency}.
We also successfully applied this technique to the problem of
selecting the best model for the observed distribution of black-hole
X-ray binary \ilya{masses}, as described in \cite{Farr2010}.  In addition to
model comparison, the PDF interpolation described here can be useful
in multi-step MCMCs to \sout{improve sampling efficiency by selecting
better jump proposal distributions} \ilya{propose jumps that can sample the parameter space more efficiently} or to test MCMC convergence. \ilya{[What are multi-step MCMCs (first mention)?  How do you use interpolation to test convergence?  I think it's bad form to introduce new ideas or new terminology in the penultimate paragraph of the paper with zero explanation.]}

>>>>

In the model selection example from Section \ref{sec:efficiency}, the
models have two-dimensional parameter spaces; in \cite{Farr2010} the
highest-dimensional model had five dimensions.  As the number of
dimensions increases the interpolation becomes more difficult for two
reasons.  First, the number of samples required for a given
number of subdivisions in each dimension grows exponentially with
the number of dimensions; hence, for a reasonable number of samples, a high-dimensional kD tree will have few subdivisions along each dimension.  Second, as
the dimensionality increases, the fraction of the volume at the ``edges'' and ``corners'' of each kD-tree box becomes more significant, and the placement of the sample becomes increasingly unrepresentative of the typical density in the box.  This problem is even more
pronounced when the distribution of samples does not align with the
coordinate axes along which the tree subdivides.

In this section, we propose a practical solution that allows us to incorporate larger-scale features in the sample density distribution.  We illustrate our solution with an single-model MCMC that updates the jump proposal distribution on-the-fly by using a kD tree to interpolate the density of existing samples.  We conclude with an example of a successful application to gravitational-wave parameter estimation.

The kD-tree based interpolated jump proposal described in Section \ref{sec:kDTree} selects one sample from the tree at random and proposes a uniform point from the box containing that sample.  This ignores the density of other samples in the neighborhood of this box, which contains additional information about the true density within the box, which is likely non-uniform.  A better proposal, therefore, would account for larger-scale sample covariances in parameter space.  To address significant covariance between samples, which would correspond to strong density gradients and a very non-uniform density profile within a rectangular box, the modified jump proposal does not descend to the leaf-nodes of the tree, but instead stops descending whenever the number of samples in the current box, $N_{box}$, falls below some threshold, $N_{crit}$. We then find the principal axes of the samples
contained in this box by calculating the eigenvectors of the covariance matrix of these samples. We use these eigenvectors to draw a new box. This box -- henceforth called a covariance cell -- contains all $N_{box}$ samples from the kD-tree box, is centered on the mean of these samples, and has edges that are aligned with the principal axes. Furthermore, this box is drawn around the samples as tightly as possible; in other words, for each edge, there exists a point in the covariance cell that lies along that edge. Figure \ref{fig:PCC} illustrates these two boxes.

\begin{figure}
  \includegraphics[width=0.8\columnwidth]{PCC}
  \caption{\label{fig:PCC} A two-dimensional illustration of the two
    boxes involved in the modified kD proposal.  The larger box
    aligned with the axes is the normal kD-tree box containing the
    given samples, while the tighter box is the ``covariance cell'' aligned with the eigenvectors of the sample distribution.  The modified kd-interpolated jump proposal draws a point from the intersection of the two boxes. }
\end{figure}

The jump proposal chooses a stored sample at random and finds the largest kD-tree box containing it and fewer than $N_{crit}$ total samples. It then draws the covariance cell and proposes a new point uniformly from the intersection of the covariance cell with the original kD-tree box.
The jump probability for this jump proposal is given by
%
\bel{modforward} P(\vtheta_i \rightarrow \vtheta_{i+1}) =
\frac{\Nbox}{N V} , \ee
%
where $N$ is the total number of samples in the kD tree, and $V$ is the
volume of the intersection between the kD-tree box and the correlation cell containing $\vtheta_{i+1}$. When used as the only jump proposal for an
RJMCMC, it is important that this proposal be capable of
proposing points in all allowed regions of parameter space; on the
other hand, when the kD proposal is used in a suite of other jump proposals, this is no longer a requirement.

@@

This technique trivially violates detailed balance since the jump proposal distribution depends on the past history. One way to address this problem requires continuously accumulating samples into the kD tree as the MCMC explores more
of the parameter space. As the number of samples drawn from the posterior PDF increases, the neighborhoods around each sample decrease with increasing sample density and the calculated covariances between samples in the kD tree better approximate the true covariances between parameters in the equilibrium posterior PDF.  Hence, for large $N$, the change in $P(\vtheta^{(j)} \rightarrow \vtheta^{(j+1)})$ approaches zero and the jump proposal does a progressively better job of sampling the equilibrium posterior while becoming asymptotically Markovian.  Conservatively, the entire set of samples obtained while the jump proposal is being dynamically updated could be discarded as burn-in, and future analysis could use the static jump proposal distribution interpolated from samples accumulated during the extended burn-in phase. 
%Because this introduces a dependence on the
%previous states of the MCMC, the jump proposal is no longer
%Markovian.  However, it remains \emph{asymptotically} Markovian \ilya{[And being asymptotically Markovian (whatever that actually means -- at least a reference?) -- is provably sufficient to guarantee that we sample from the equilibrium distribution even though we nominally violate detailed balance? (again, a reference?)]} , which
%ensures that the equilibrium distribution of the MCMC is correct.

To efficiently insert samples into the tree as the chain accumulates them,
the algorithm from Section \ref{sec:kDTree} must be modified:
\begin{enumerate}
\item Instead of partitioning the sample set at its median, we now
  partition the bounding box at its geometric center along a
  particular dimension, which cycles as we descend the tree levels.
  Note that this allows for empty boxes if all the samples cluster to
  one side of the domain.
\item When inserting a sample, we descend the tree to a leaf node,
  which now can be either empty or contain one sample.  If it is empty,
  we insert the sample at this node.  If it contains one sample, we
  subdivide the appropriate dimension, and place both samples into the
  two sub-boxes.  If both land in the same sub-box, we continue
  subdividing, until each box in the sub-sub-sub... tree contains one
  or zero samples.
\end{enumerate}
%Insertion of a sample is $\order{\log N}$, since it involves a constant amount of work per tree level, and there are $\log N$ tree levels.  \ilya{[No -- the number of tree levels can be arbitrarily large if the original samples are clustered into a very small region of the prior volume...]} \dan{[Is this true? Couldn't it be arbitrarily small, depending on how far into the tree you want to burrow (i.e. depending on $N_{crit}$)?]} Neighborhood lookup proceeds as before, and is therefore also $\order{\log N}$.  \ilya{[Perhaps $\order{\log N + \Nbox}$, since must recompile the inner bounding box every time -- and, depending on problem, $\Nbox$ could be $\gg \log N$?]} \dan{[Need only recompute the covariance cell if a new sample was accepted into the corresponding kD tree cell between calls to that kD tree cell. I'm not sure how that would factor in here, but I feel like it should.] }

We have implemented this modified kD tree proposal as one of many jump
proposals in the \texttt{LALInferenceMCMC} sampler
\cite{vanderSluys:2008a,Raymond:2010,Raymond2012,Veitch:2014}.
\texttt{LALInferenceMCMC} is a MCMC code, based on the LIGO algorithms
library ({\tt http://www.lsc-group.phys.uwm.edu/lal}), designed to sample the posterior on parameters of merging
compact-object binaries (masses, sky location, orbital orientation, distance, etc.) 
encoded in their gravitational-wave signatures as observed by the ground-based
gravitational-wave detectors LIGO \cite{AdvLIGO} and Virgo \cite{AdvVirgo}.  The simplest such
signal has a nine-dimensional parameter space, and the posterior often
includes near-degeneracies and multiple modes that make convergence of
the MCMC chain very slow with traditional proposals \cite{S6PE}.

Figure \ref{fig:accratio} shows the acceptance ratio for the kD-tree jump
proposal applied as part of a \texttt{LALInferenceMCMC} analysis.  In this example, the kD-tree jump proposal is one of several jump proposals used during the MCMC, and the kD tree itself is updated with accepted samples whenever its jump proposal is called. In spite of the very low acceptance
rate of the proposal, applying the kD-tree proposal to one out of
20 proposed jumps improved the convergence time of the simulation
by a factor of two compared to the standard suite of proposals because
it is particularly efficient at producing ``mode-hopping'' jumps,
which are difficult to produce with the other proposals in
\texttt{LALInferenceMCMC}.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{linearlog3}
  \end{center}
  \caption{\label{fig:accratio} The cumulative parameter acceptance
    ratio for the modified kD-tree jump proposal used in the
    \texttt{LALInferenceMCMC} code as a function of the number of
    steps (in hundreds). In spite of the small acceptance ratio, the
    kD tree proposal, applied to one in 20 proposed jumps,
    improved the convergence time of the sampler by a factor of two
    compared to using only the standard suite of proposals. 
    The acceptance rate asymptotes to the steady-state solution once sufficient samples have been accumulated in the kD tree to allow the sample density to be accurately interpolated.
%\dan{[Taking a leap here, someone please confirm/deny/add qualifiers] The initial decrease in acceptances arises from insufficient population of the kD tree with too few samples, while the leveling-off later in the MCMC indicates convergence.} \ilya{[I see structure in this figure (a high acceptance probability, followed by a dip, followed by an asymptote).  Is this a fluke, or do you always see this -- and, if so, why?  BTW, what's the autocorrelation length (just to give me a sense of scale, so I am not reading too much into meaningless fluctuations within an AC length)?]} \dan{[Don't recall the autocorrelation length. Will?]}
}
\end{figure}

@@

The need to compare evidences for multiple models arises in a large
variety of physical and astronomical contexts.  In this paper, we
described a technique that allows for efficient evidence
computations via a Reversible-Jump Markov chain Monte Carlo.  This
technique solves the usual problem of finding good inter-model jump
proposals in an RJMCMC by using a kD tree to quickly and accurately
interpolate an approximate posterior PDF from a single-model MCMC run,
and then proposing efficient inter-model
jumps from this interpolated PDF.

We demonstrated the efficiency of this technique on a toy
model-comparison problem described in Section \ref{sec:efficiency}.
We also successfully applied this technique to the problem of
selecting the best model for the observed distribution of black-hole
X-ray binary masses, as described in \cite{Farr2010}.  In addition to
model comparison, the PDF interpolation described here can be useful
in single-model MCMCs to inform the jump proposal distribution on the fly in order to propose jumps that can efficiently sample the parameter space (see Section \ref{sec:higherdim}), or to test MCMC convergence.
